---
title: "The Procedural Flaws of the Chilean Constitutional Process (2019-2022)"
author: "María Isabel Álvarez"
date: "March 7, 2023"
output: pdf_document
bibliography: ChileanCCReferences.bib
csl: chicago-author-date.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("papaja")
r_refs("r-references.bib")
library(readxl)
library(knitr)
opts_template$set(figure1 = list(fig.height = 6, fig.width = 12), 
                  figure2 = list(fig.height = 8, fig.width = 15),
                  figure3 = list(fig.height = 6, fig.width = 12),
                  figure4 = list(fig.height = 6, fig.width = 12),
                  figure5 = list(fig.height = 6, fig.width = 12),
                  figure6 = list(fig.height = 6, fig.width = 12),
                  figure7 = list(fig.height = 6, fig.width = 12),
                  figure8 = list(fig.height = 7, fig.width = 10))
library(tidyverse)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")
CCRollCallVotesRegDelegateVotes <- read_excel("CCRollCallVotesRegDelegateVotes.xlsx")
CCRollCallVotesSubsDelegateVotes <- read_excel("CCRollCallVotesSubsDelegateVotes.xlsx")
```


# Introduction
The topic that I will be addressing in my BA/MA thesis is the failure of the 2019-2022 Chilean Constitutional Process. The project to rewrite a constitution to replace the Constitution ratified in 1980 during General Augusto Pinochet’s military dictatorship (and amended substantively in 2005 by former-President Ricardo Lagos) had massive initial popular support after the 2019 Chilean Estallido Social (Social Outbreak) protests. To be precise, the referendum that formally began the process of Chile’s constitution-making had 78.28% of approval (with a 50.95% of voter participation) and a Constitutional Convention was chosen as the preferred organ to redraft the constitution with 79% approval [@servel2020] ^[The plebiscite had two questions, whether Chileans approved or rejected the redrafting of a new Constitution and if so, which organ would be charged with the drafting of the text, either a Mixed Constitutional Convention (made up 1/2 by parliamentarians and 1/ of elected members) or a Cosntitutional Convention (made up 100% of elected members)] . Two years later, a referendum was held to reject or ratify the Constitutional Proposal draft; and the results were loud and clear: the Proposal was rejected by 61.89% of the votes, and this election broke a record of voter participation in Chile with an 85.86% of participation [@servel2022]. In other words, Chileans expressed “we want a new Constitution, but certainly not this one”.

Putting these results in a historical and comparative perspective, out of the 179 plebiscites on new constitutional processes that have taken place around the world between 1789 and 2016, only 6% of these have rejected a newly drafted constitution [@elkins2019constitutional]. The failure of Chile’s constitutional process was as historic as was the opportunity to write a constitution under the conditions that Chilean Constitutional Convention drafted it: with gender parity, regional representation, and a proportional Indigenous peoples’ quota. 

What went wrong for this project to have such a fall from grace? Politicians, academics, and civil society have put out there their own arguments and diagnoses on the causes for the failure of the Constitutional Process. And this thesis project aims to research one of the wildly overlooked causes of failure: the procedural components of the Process. 

The research questions I am posing for this thesis project are: To what extent can the procedural framework of the Constitutional Convention be considered one of the main causes for the failure of the Constitutional Process? What features of the procedural and regulatory architecture in the Chilean Constitutional Process hindered collective wisdom and rational decision-making? What are the implications of not achieving democratic rationality in a Constitutional Process? But considering the narrower framework of this report, I will be focusing on the (externally impose) procedural mechanisms that aimed to broaden the inclusivity of the Constitutional Convention. I will analyze, given these procedures, whether cognitive diversity (a fundamental condition for democratic reason) was achieved within the Constitutional Convention or not. And after determining if it was achieved or not, what are the implications of cognitive diversity having been (or not) achieved in the Constitutional Convention?  What are the implications for party dynamics?


# The Argument and Relevance to the Data
The Chilean Constitutional Convention was bound by a wide array of procedures which were developed in different stages of the Constitutional Process. Some of these procedures were developed endogenously, meaning certain procedures were drafted and adopted, by and for the Constitutional Convention itself. But some of these procedures were developed exogenously, as institutions and bodies external to the Constitutional Convention drafted and imposed mechanisms to constrain the Convention before the Convention convened for the first time.  

The argument I pose is that the Chilean Constitutional Process failed greatly due to the exogenously imposed and endogenously adopted procedures that the Constitutional Convention was bound by. The Constitutional Convention (CC), the body charged with the drafting of the Constitutional Draft Proposal made up of elected citizen-legislators, was bound by four sets of procedural norms– three externally developed and one internally developed. Without getting into too much detail, I argue that the Constitutional Convention failed due to its externally and internally designed procedural architecture insofar as both hindered the achievement of democratic reason in the Convention’s decision-making processes. This argument will evaluate the impact of procedural features on the decision-making processes and dynamics of the Constitutional Convention from a political epistemology perspective. 

My central hypothesis is that the procedural frameworks– both, those imposed on by a body external to the Convention and those internally adopted by the Convention– severely hindered the Constitutional Process as a whole and can be considered a determinant cause for its failure. 

This report will particularly focus on the argument that despite there being positive externally-imposed procedures that not only aimed but succeeded at broadening inclusivity within the Constitutional Convention, the process still failed. The Chilean Congress purposefully passed a number of laws that broadened the levels of inclusivity and increase the levels of representation [@cembrano2021propapp]. The three main ones considered in this report are: gender parity, Indigenous peoples' quotas and the participation of Independents [@ley21298; @ley21216]. Each of the three procedures (legislated in two laws) had a monumental impact on the Convention. And the two categories of analysis, the two main impacts evaluated in this report are how these procedures intended to broaden inclusivity within the Convention affected 1) Cognitive Diversity, and 2) Party Unity within the Constitutional Convention. 

The argument is relevant because inclusivity and high levels of representation are (rightfully) praised as valuable and even necessary to democratic processes. And the most ambitious arguments, as I will outline in a following section, present the case that cognitive diversity is not only a prerequisite to achieve democratic reason but also that group diversity is more important than the individual ability of the members in a decision-making body in democratic processes [@delcogdiv; @landemore2012collective]. This report will show that broad inclusivity was achieved, and that cognitive diversity was achieved, too. But it will also show that broad inclusivity within the Constitutional Convention was not only not enough for the Constitutional Process to succeed, but that the broad diversity at times hindered democratic reason insofar as it hindered meaningful intersectional collaboration and consensus. 
	
# Presentation of the Data
This scientific report will consider a specific part of this argument, which is how the procedures affected the dynamics of the Constitutional Convention. The data considered in the scope of this report only considers the Constitutional Convention– its makeup and its dynamics. 

The first dataset is qualitative as it contains the Constitutional Convention delegate information. Here, there is information about each delegate’s demographic background, their election in the convention, their political affiliation and their educational and professional careers. This qualitative dataset will be used to analyze cognitive diversity. 

The second and third datasets contain all of the roll call votes casted in the Constitutional Convention (divided by phase– Regulatory and Substantive). The second dataset is quantitative as it contains the votes casted in the Regulatory Phase by the Constitutional Convention delegates. The regulatory phase is composed of 1009 roll call votes in which the Constitutional Convention delegates produced their own regulatory handbook or bylaws. The third data is quantitative, too, as it contains the votes casted in the Substantive Phase by the Constitutional Convention delegates. And the substantive phase is composed of 3508 roll call votes in which the Constitutional Convention delegates determined the content of the Constitutional Proposal Draft. These two datasets will be analyzed in regards to party unity, focusing primarily on the patterns by political party and by political tendency. Analyzing these frequency distributions and standard deviations is valuable to evaluate the consistency of voting patterns and unity within political parties in the Convention. 

# Cognitive Diversity 
Landemore has built a formidable case for the epistemic value and virtue of inclusive deliberative democracy that is based on the cognitive diversity of the group engaged in collective decision-making [@landemore2017democratic]. Democratic reason, the collective distributed intelligence of the people, can be achieved through certain mechanisms which are political cognitive artifacts; the two main being deliberation and majority rule. Both have their own specific epistemic properties and values. But this section will focus on the concept of cognitive diversity, which is considered a fundamental condition/characteristic conducive to democratic reason. 

Cognitive diversity refers to the “the variety of mental tools that human beings use to solve problems or make decisions in the world” [@delcogdiv, pp. 89]– in other words, a mental toolkit that is the product of a unique living experience. It is theoretically considered to be the condition of optimal deliberation; meaning it is foundational to the concept of democratic reason and its mechanisms of deliberation and majority rule. This is the condition of optimal deliberation because it dictates that cognitive diversity in a group matters more than individual epistemic competence. And cognitive diversity brings with it democratic reason. 

And this section will analyze if that particular argument is supported or not. Was the Constitutional Convention a decision-making body with cognitive diversity? As aforementioned, several procedures that were externally imposed by the Chilean Congress were designed specifically to broaden the inclusivity of the Convention (gender parity, indigenous peoples’ quotas and the participation of Independents). These were celebrated measures and the measures that made the Chilean Constitutional Process so unique– there has been no constitution-making organ that has drafted a constitution under these conditions. But how did broad inclusivity impact the Constitutional Process in regards to facilitating cognitive diversity? 

## Data Analysis with Delegate Information Dataset
Adding to the aforementioned definition, "cognitive diversity “refers to a diversity of ways of seeing the world, interpreting problems in it and working out solutions to these problems” and “denotes more specifically a diversity of perspectives (ways of representing situations and problems), diversity of interpretations (ways of categorizing or partitioning perspectives), diversity of heuristics (ways of generating solutions to problems) and diversity of predictive models (ways of inferring cause and effect)”[@delcogdiv]. But cognitive diversity is “conceptually distinct from both some of its causes (e.g., gender, ethnicity, or, more fundamentally, genes) and some of its symptoms (e.g. differences in viewpoints or opinions”– meaning cognitive diversity goes beyond broad representation of different groups but gets to a diversity in cognitive processes. In other words,  “The diversity that really matters is not primarily a diversity of opinions, values, perspectives (as end-results rather than processes), or even a diversity of social and economic backgrounds. What matters is a more fundamental cognitive diversity defined as the internal, psychological property that determines how each individual sees the world, interprets its problems and maks predictions in it.” The importance of cognitive diversity and the argument that Landemore, alongside Lu Hong and Scott Page make, is that the cognitive diversity in a group is more valuable than the average individual ability for a group’s collective competence in decision-making– particularly, democratic decision-making.  

I will argue that the data presented below does demonstrate, to a certain extent, that there was an achievement of cognitive diversity in the Constitutional Convention. Cognitive diversity is not a measurable or quantifiable concept. And while the theorists argue against equating cognitive diversity with its symptoms or causes, I will at at least minimally be able to argue that there were the optimal conditions for there to be cognitive diversity within the Constitutional Convention. An maximally I will argue that the blend of these very different Chilean experiences did result in cognitive diversity within the Convention. 

Turning to the data, Tables 1 and 2 summarize the frequency of a number of demographic variables which demonstrate diversity within the Constitutional Convention at a very basic level. These tables show the frequency of the delegate’s gender, age range, and macrozone. This basic demographic information shows the distribution of the Constitutional Delegate’s gender, age group and general geographical origin. The data is relevant because it shows that the Constitutional Convention, at a basic level had gender parity, representation of different age groups and generations, and complete geographical representation of Chile. Table 1 shows the relationship between age range and gender within the convention. Gender was equally represented in the Convention (77 members were male-identifying and 77 members were female-identifying) and all in all, there is vast generational representation. But the age group with the broadest representation was 30 to 39 years old, and the age group with the smallest representation was 20 to 29. Furthermore, Table 2 shows the macrozone representation. Chile has 16 regions that extend over the longest country in the world of length, but its government and its institutions are highly centralized. Which is why macrozone representation is important to be included in its diversity. While the most represented area is its Metropolitan Region (the capital, Santiago), each macrozone (except the Austral) has around 20 representatives. 

```{r table-1, echo= FALSE}
#Read in the data + viewing the data 
library(readxl)
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")

#Table and Kable showing gender and age range distribution 
genderage <-  table(ConstitutionalConventionDelegateInfo$Gender, 
                    ConstitutionalConventionDelegateInfo$`Age Range`)
knitr::kable(head(genderage[, 1:5]), "pipe", caption = "Table showing the frequency of age ranges by gender in the Constitutionional Convention with gender parity (and a total of 77 female-identifying delegates and 77 male-identifying delegates")

```


```{r table-2, echo= FALSE}
#Table and Kable showing gender and age range distribution 
gendermacrozone <-  table(ConstitutionalConventionDelegateInfo$Gender, 
                    ConstitutionalConventionDelegateInfo$Macrozone)

knitr::kable(head(gendermacrozone[, 1:6]), "pipe", caption = "Table showing the macrozone distribution in the Convention by gender. Austral= Austral Zone, Centro= Cental Zone (excluding the Metropolitan zone), Centro Sur= Central Southern Zone, Metropolitana= Metropolitan Zone (Santiago, the Chilean Capital, Norte Grande= Northern Zone, Sur= Southern Zone")

```

Turning to figures, these show a more in-depth level of cognitive diversity through the different academic and professional backgrounds of the Constitutional Convention delegates. There was a variety of educational levels represented in the Convention, as well as professions. This is data analysis is important because it essentially supports the argument that the Convention was not just a group of technocrats or just a group of lawyers and politicians drafting a new Constitution. It is observable that several backgrounds that are accompanied by different skill-sets alongside different problem-diagnosis-and-solving abilities were represented. 

Figure 1 displays the variety of educational levels represented in the Convention. Diversity in educational levels range from a completed elementary school education to an array of graduate degrees, and everything in between. The representation of a wide array of educational experiences is valuable to broaden inclusivity, especially in a foundational democratic process like constitution-making. Different experiences yield different perspectives, which consequently yield different approaches to problem identification and solving, i.e. cognitive diversity. Not to mention, different educational backgrounds will provide valuable input in institution-making. 

```{r figure-1, opts.label = "figure1", echo=FALSE, fig.cap= "Constitutional Convention Delegates and their Completed/Current Educational Level"}
library(readxl)
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")

#Plot showing the CC makeup by the delegate's completed educational level 
ConstitutionalConventionDelegateInfo %>% 
  group_by(lastname_firstname, educational_level) %>%
  ggplot(aes(x =educational_level, fill= educational_level)) +
  theme(plot.title = element_text(face = "bold")) + 
  geom_bar() +
  labs(y = "Frequency", x = "Completed/Current Educational Level") +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +
  geom_text(aes(label = after_stat(count)), stat = "count", vjust = 1.5, colour = "white")

```

Figure 2 displays the variety of professions represented in the Convention. And this professional diversity yields very similar inputs as educational diversity. And I argue that this information truly is the key to unlocking the argument that there was cognitive diversity in the Convention given the vast array of cognitive processes represented in the constitution-making body. 

Constitution-making is an incredibly complex and technical process. Technocratic arguments already favor technical expertise over representation and inclusivity. But these arguments are doubled on specifically when applied to constitution-making given the intricacies and legal expertise "required" to partake in such processes. Nonetheless, technical expertise is not always attached or related with educational experience, which is why either way it is helpful to look at professions. But Landemore argues that cognitive diversity can and will make up for lack of technical expertise. Cognitive diversity on its own is already valuable and conducive to democratic reason, but cognitive diversity can also make up for technical inexpertise or knowledge on a subject matter. 

```{r figure-2, opts.label = "figure2", echo=FALSE, fig.cap="Constitutional Convention Delegates and their Profession"}
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")

#Plot showing the CC makeup by the delegate's profession 
ConstitutionalConventionDelegateInfo %>% 
  group_by(lastname_firstname, profession) %>%
  ggplot(aes(x = profession, fill= profession)) + geom_bar() +
  theme(plot.title = element_text(face = "bold")) + 
  labs(y = "Frequency", x = "Profession") +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +
  geom_text(aes(label = after_stat(count)), stat = "count", vjust = 1.5, colour = "white")
```


Without the procedures that the Chilean Congress designed or imposed onto the Convention, this degree of representation (and therefore cognitive diversity) would have not been achieved. Regional representation would have been achieved given the default on electoral laws; but it is highly doubtful that the electorate would have been able to elect a perfectly equal distribution of female and male identifying delegates. And most definitely, this degree of Indigenous peoples’ representation would have not been possible without these explicit pieces of legislation that automatically reserved 17 seats out of the 155 seats in the Convention for ten different groups [@ley21298]. Furthermore, the facilitated participation of Independents, arguably, is what most influenced the conditions for there to be cognitive diversity in the Convention [@ley21216]. As Independents are people not formally affiliated with politicians, these are people who have most definitely not been formally involved in politics. The values and limitations of this are reviewed at length in my thesis project. But focusing on this report’s aim, by actively involving people that do not have a political background or profession in a political process, lots of non-political backgrounds and professions were represented, backgrounds and professions that require different skill sets, perspectives and problem-solving abilities. 

# Party Dynamics 
This section also considers the impact of the externally imposed procedure of the participation of Independents on another variable, party dynamics. The Chilean Congress passed specific laws that first allowed the participation of Independents in the election of the Constitutional Convention [@ley21216] delegates and then passed another law that would facilitate their inscription process[@ley21296]. In other words, Independents were incentivized to participate. In brief, this was a reflection of the Congress appeasing to the anti-party mood in Chile after the Social Outbreak protests and the will of the people to draft a new Constitution by and for “the people”– which is why a Constitutional Convention was elected as the constitution-making organ instead of a Mixed Constitutional Convention in which half of the body would be made up of elected citizens and half of current parliamentarians. Chileans, by electing with an overwhelming electoral majority a Constitution Convention whose all members had to be elected citizens, demanded to be rid of traditional party-politics for this process [@servel2020]. 

Figure 3 shows the distribution of the general political parties represented. I specify that this is the general political party distribution because  most delegates ran under very small Independent lists and Indigenous peoples groups ran under their particular peoples groups (Pueblos Originarios). In total and in detail, there technically were 51 different lists/parties represented among 154 delegates. But when conflating all of the Independents in one category, and the Indigenous peoples quotas into on category, we’re left with 12 “political parties”– 10 of these are established Chilean political parties, one contains all delegates that ran as Independents and one contains all the different Indigenous delegates representing their peoples. As Figure 5 shows, Independents overwhelmingly hold the concentrated majority. A total of 87 of the 154 delegates ran through Independent lists, and the second largest concentration are the 17 Indigenous peoples’ reserved seats. After that, three established political parties were tied at 10 delegates.  

```{r figure-3, opts.label = "figure3", echo=FALSE, fig.cap="Distribution of Political Parties, Independents and Indigenous peoples' quotas in the Constitutional Convention"}
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")

#Plot showing the general (concept explained in the 
#manuscript) frequency of the  political parties + independents in the Constitutional Convention 
ConstitutionalConventionDelegateInfo %>%
  group_by(lastname_firstname, polparty_gen) %>%
  ggplot(aes(x= polparty_gen, fill = polparty_gen)) + geom_bar() +
  labs(y = "Frequency", x = "Political Party") +
  theme(plot.title = element_text(face = "bold")) + 
  theme(axis.text.x = element_text(angle = 65, vjust = 0.9, hjust=1))+
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white")

```


Figure 4 shows the distribution of the different political tendencies represented. Again, I have grouped Independents into a category and Indigenous peoples into another category. This is to primarily focus on the distribution of ideologies from the traditional political paries in Chile. Oficialismo refers to the governing coalition– considered here as current-President Gabriel Boric’s coalition which encompasses center to left wing parties. And Oposicion refers to the opposing coalition– considered here as the current opposition of center to right wing parties. One of the central procedures that my thesis project focuses on is the 2/3 supermajority voting quorum needed to pass any norm in the substantive phase of the Process (substantive meaning that what passed, made it onto the Constitutional Draft). And it is because given that the opposition (center to right wing parties) were not able to achieve a third of the seats in the Constitutional Convention, the Chilean right was left with no bargaining power in agreements as they didn't have an informal "veto" power or the numbers to block motions [@interferencia1]. As a result, I argue, an entire sector of the Chilean political spectrum was ignored in negotiations; they did not have to be taken seriously because they did not have the numbers to be meaningfully included in negotiations. This, was counterproductive because it produced an environment of adversarial instead of consensus politics; delegates from different ideological currents saw each other as barriers and obstacles to their own goals instead of fellow collaborators, all because "what really mattered" was achieving a 2/3 supermajority [@interferencia1]. 

```{r figure-4,opts.label = "figure4", echo=FALSE, fig.cap="Distribution of Political Tendencies in the Constitutional Convention (Independientes = Independents, Oficialismo= Government, Oposicion= Opossition, Pueblos Originarios = Indigenous peoples' quotas)"}
ConstitutionalConventionDelegateInfo <- read_excel("ConstitutionalConventionDelegateInfo.xlsx")

#Plot showing the CC makeup by political tendency where Independientes = Independents, Oficialismo= Government, Oposicion= Opossition, 
#Pueblos Originarios = Indigenous peoples' quotas
ConstitutionalConventionDelegateInfo %>% 
  group_by(lastname_firstname, tendency) %>%
  ggplot(aes(x = tendency, fill= tendency))  + geom_bar() + 
   labs(y = "Frequency", x = "Political Tendency") +
  theme(plot.title = element_text(face = "bold")) + 
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white")
```


## Data Analysis with Regulatory Phase and Substantive Phase Roll Call Data  
Having laid out the general political party and political tendency distribution, now I will analyze the specific voting behavior of these different political parties and political tendencies. Figure 7 and 8 show the frequency distribution in the regulatory and substantive phase of votes casted in favor by members of different political parties. 

Figure 5 shows the frequency distribution by political party regarding the in favor votes casted by the Constitutional Convention delegates in the Regulatory Phase of the Convention. Visually, the distributions are relatively similar in the sense that the average for each seem to be around the 60-70% of votes casted in favor. Not even getting into actual statistics, just visuals, I aim to show that political parties in the regulatory phase acted very similarly. Despite ideological differences, their voting patterns show there was not much internal divergences within political parties nor substantial differences relative to other parties. When drafting the bylaws, political parties according to their voting behavior, acted similarly internally and compared to each other. 

```{r figure-5, opts.label = "figure5", echo=FALSE, fig.cap= "Frequency Distribution (by Political Party) of Votes In Favor Casted in the Regulatory Phase"}
CCRollCallVotesRegDelegateVotes <- read_excel("CCRollCallVotesRegDelegateVotes.xlsx")

#FIGURE 5
#Plot showing the percentage of votes casted, polparty by frequency 
CCRollCallVotesRegDelegateVotes %>%
  group_by(polparty_gen, percentage_votesinfavor) %>%
  ggplot(aes(x=polparty_gen, y=percentage_votesinfavor, fill= polparty_gen, color= polparty_gen)) + geom_count() +
  theme(plot.title = element_text(face = "bold")) + 
  labs(y = "Percentage Votes Casted In Favor", x = "Political Party") +
  scale_y_continuous(limits = c(30,80))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
```

But Figure 6, in contrast, shows a completely different voting pattern and behavior. During the substantive phase, when the votes determined what kind of content, policies and norms went into the Constitutional Draft, parties began establishing very distinct behavior to each other. Independents have the biggest widespread voting frequency, meaning Independent members voted differently to each other. It makes sense that they did not demonstrate party unity as they were not an established party, they were Independents. On the other hand, there is radical contrast between the party unity of Independents and the party unity of established parties. Other than that, the Chilean Socialist Party was the other political grouping that did not display much party unity.

```{r figure-6,opts.label = "figure6", echo=FALSE, fig.cap="Frequency Distribution (by Political Party) of Votes In Favor Casted in the Substantive Phase"}
CCRollCallVotesSubsDelegateVotes <- read_excel("CCRollCallVotesSubsDelegateVotes.xlsx")
#FIGURE 10 
#Plot showing the percentage of votes casted in the Substantive Phase, polparty by frequency 
CCRollCallVotesSubsDelegateVotes %>%
  group_by(polparty_gen, percentage_infavor) %>%
  ggplot(aes(x=polparty_gen, y=percentage_infavor, fill= polparty_gen, color= polparty_gen)) + geom_count() +
  labs(y = "Percentage of Votes Casted In Favor", x = "Political Party") +
  theme(plot.title = element_text(face = "bold")) + 
  scale_y_continuous(limits = c(20,80))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Figure 7 and 8 show the frequency distribution in the regulatory and substantive phase of votes casted in favor by members of different political tendencies. Boiling down groupings by political trends allows for a more digestible interpretation for those who aren't familiar with the Chilean political landscape and spectrum. While it is a simplification and that can be a limitation, it is a useful way to process the information. Figure 7 shows that there was not much tendency unity during the regulatory phase for the debate for the Opposition coalition and the Indigenous peoples' groups. But that there was tendency unity for Independents and the Government coalition. While the Government coalition does have outliers in both extremes, delegates that voted in favor of a lot of norms and delegates that voted against a lot of norms, it still displays unity within the political trend. 
 

```{r figure-7, opts.label = "figure7", echo=FALSE, fig.cap= "Frequency Distribution (by Political Trend) of Votes In Favor Casted in the Regulatory Phase"}
CCRollCallVotesRegDelegateVotes <- read_excel("CCRollCallVotesRegDelegateVotes.xlsx")

#FIGURE 7
#Plot showing the percentage of votes casted, political trend by frequency 
CCRollCallVotesRegDelegateVotes %>%
  group_by(tendency, percentage_votesinfavor) %>%
  ggplot(aes(x=tendency, y=percentage_votesinfavor, fill= tendency, color= tendency)) + geom_count() +
  theme(plot.title = element_text(face = "bold")) + 
  labs(y = "Percentage Votes Casted In Favor", x = "Political Party") +
  scale_y_continuous(limits = c(40,80))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```
Table 3 and Figure 8 show the frequency, distribution and deviation in the Substantive phase by political tendency. Table 10 numerically shows via the calculation of standard deviations what is visually displayed in Figure 11. Table 10 and Figure 11 show that the Independents had the biggest deviation; which is unsurprising given their whole point was not to behave like a political party. But surprisingly (or not), the second largest deviation is the Government's party coalition. The Chilean center to left parties very publicly became ideologically fragmented, but this data confirms it numerically and contained in this constitution-making body. The opposition coalition, the center to right wing parties displayed a decreased standard deviation by around 35%, showing that they remained relatively united throughout the Substantive phase according to their voting behavior. But the political tendency with the smallest standard deviation and the narrowest frequency distribution is the Indigenous peoples' groups. Their voting patterns remained relatively consistent within that political trend. But also a potential confounder variable could be that their standard deviation is lower because they are also the smallest group with only 17 members. Nonetheless, these voting behaviors show that in the Convention, Independents and the Government coalition lacked unity within their political tendency, while the Opposition coalition displayed more unity but the most united political tendency were the Indigenous peoples' groups. 

```{r table-3, opts.label = "figure10", echo=FALSE}
#STANDARD DEVIATION BY TENDENCY 
sdpoltend <- CCRollCallVotesSubsDelegateVotes%>%
  group_by(tendency) %>%
  summarise_at(vars(percentage_infavor), list(name=sd))

#Table SD by tendency in knitr format 
knitr::kable(head(sdpoltend[, 1:2]), "pipe", col.names = c("Political Tendency","Standard Deviation"), align = "c", digits = 2, caption = "Table showing the standard deviation by political tendency in the Substantive phase of the votes. Independientes = Independents, Oposicion = Opposition coalition, Oficialismo= Government coalition, Pueblos Originarios = Indigenous peoples' groups")

```

```{r figure-8, opts.label = "figure8", echo=FALSE, fig.cap= "Frequency Distribution (by Political Tendency) of Votes In Favor Casted in the Substantive Phase"}
CCRollCallVotesSubsDelegateVotes <- read_excel("CCRollCallVotesSubsDelegateVotes.xlsx")

#FIGURE 8
#Plot showing the percentage of votes casted in the Substantive Phase, poltend by frequency 
CCRollCallVotesSubsDelegateVotes %>%
  group_by(tendency, percentage_infavor) %>%
  ggplot(aes(x=tendency, y=percentage_infavor, fill= tendency, color= tendency)) + geom_count() +
  theme(plot.title = element_text(face = "bold")) + 
  labs(y = "Percentage of Votes Casted In Favor", x = "Political Tendency") +
  scale_y_continuous(limits = c(20,80))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

```


# Conclusion
Thus, this report has evaluated the impact of some externally imposed procedures that aimed to broaden the Constitutional Convention’s inclusivity on the Convention’s internal dynamics focusing on cognitive diversity and party unity. Inclusivity and representation are rightfully celebrated in democratic processes. Theorists like Landemore, Ho and Page show how inclusivity and representation can be conducive to cognitive diversity in a democratic process. They argue that diversity in a decision-making group is more valuable than the average individual ability for a group’s collective competence particularly in democratic processes. But the Chilean Constitutional Process seems to be an outlier. Cognitive diversity was not enough to guarantee the Process' success and it seemed to hinder the ability of the Convention to reach meaningful levels of consensus and agreement. While there were high degrees of representation due to procedural mechanisms that aimed to purposefully design an inclusive constitution-making organ, and while there seemed to be cognitive diversity, these features counter-intuitively hindered democratic reason. They counter-intuitively hindered democratic reason because they were not conducive to consensus even after thorough rounds of deliberation and majoritarian votes. Focusing on these majoritarian votes, it is evident that while the Convention's parties and political tendencies remained united during the regulatory phase internally and externally in comparison to other parties and tendencies, that unity seemed to decrease in the substantive phase both internally and externally in comparison for other parties and tendencies. 

# References
<div id="refs"></div>

